---
title: "Transcriptomic age from bulk RNA-seq with tAge"
author: "Gladyshev Lab"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Transcriptomic age from bulk RNA-seq with tAge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tAge)

# This vignette runs on the example data shipped with the package. The
# prediction step additionally needs a Python environment (reticulate +
# joblib/pandas/scikit-learn/numpy) and downloads a small clock model from
# Zenodo. When either is unavailable, the prediction chunks are skipped so the
# vignette still builds. To run the full pipeline locally, set
# RETICULATE_PYTHON to your environment before building:
#   Sys.setenv(RETICULATE_PYTHON = ".venv/bin/python")

.py_ok <- requireNamespace("reticulate", quietly = TRUE) && tryCatch(
  all(vapply(c("joblib", "pandas", "sklearn", "numpy"),
             reticulate::py_module_available, logical(1))),
  error = function(e) FALSE)

.model_path <- NULL
if (.py_ok) {
  .model_path <- tryCatch({
    cl <- list_clocks(type = "EN", outcome = "Mortality",
                      species = "Multispecies", tissue = "Multi-Tissue",
                      scaling = "Scaled")
    download_clocks(cl, dest_dir = file.path(tempdir(), "tage-clocks"),
                    quiet = TRUE)$path[1]
  }, error = function(e) NULL)
}

RUN_MODELS <- .py_ok && !is.null(.model_path) && file.exists(.model_path)
```

# Overview

`tAge` applies the transcriptomic clocks from
[Tyshkovskiy et al. (2026), *Nature*](https://doi.org/10.1038/s41586-026-10542-3)
to bulk RNA-seq data, predicting **chronological age** (in age units) and
**expected mortality** (as a `log10` hazard ratio).

The workflow: build an `ExpressionSet`, preprocess it, and predict.

# 1. Build an ExpressionSet

We use the mouse example data bundled with the package (genes × samples raw
counts, plus sample metadata).

```{r data}
exprs_data <- load_example_expression_data()
metadata   <- load_example_metadata()

eset <- make_ExpressionSet(exprs_data, metadata, verbose = FALSE)
eset
```

# 2. Preprocess

`tAge_preprocessing()` filters low-expressed genes, maps them to the mouse
Entrez gene space, RLE-normalises, log-transforms, scales (per sample), applies
the YuGene transform, aligns to the clock gene list, and centres the data.

```{r preprocess}
tAge_eset <- tAge_preprocessing(
  eset,
  species = "mouse",
  gene_mapping_type = "Ensembl",
  verbose = FALSE
)
names(tAge_eset)
```

Prediction uses the `scaled_diff` and `yugene_diff` representations.

# 3. Browse and download clocks

The published clocks live on Zenodo; list them from R (reads the bundled
registry):

```{r list-clocks}
head(list_clocks(type = "EN", outcome = "Mortality"))
```

`download_clocks()` fetches a chosen set and returns a `path` column:

```{r download, eval = FALSE}
clocks <- list_clocks(type = "EN", outcome = "Mortality",
                      species = "Multispecies", tissue = "Multi-Tissue")
clocks <- download_clocks(clocks, dest_dir = "clocks")

model_paths <- list(
  scaled_diff = clocks$path[clocks$scaling == "Scaled"],
  yugene_diff = clocks$path[clocks$scaling == "YuGene"]
)
```

# 4. Predict

```{r predict, eval = RUN_MODELS}
Sys.setenv(RETICULATE_PYTHON = Sys.getenv("RETICULATE_PYTHON"))  # use your env

results <- predict_tAge(
  tAge_eset,
  model_paths = list(scaled_diff = .model_path),
  species = "mouse",
  mode = "EN"
)
head(results[, c("Genotype", "Tissue", "scaled_diff_EN_tAge")])
```

```{r predict-skipped, echo = FALSE, results = "asis", eval = !RUN_MODELS}
cat("> _Prediction skipped: no Python environment (or no internet to fetch the",
    "model) was available at build time. Set `RETICULATE_PYTHON` and rebuild to",
    "run this step._\n")
```

These mortality values are `log10(hazard ratio)` — small numbers centred near
zero, **not** ages. See below.

# Interpreting the output

| Outcome | Units | Notes |
|---|---|---|
| Chronological | age (years / months) | normalised age rescaled by species max lifespan |
| Mortality | `log10(hazard ratio)` | **not** an age; higher = higher expected mortality |
| Normalized age | fraction of max lifespan | native scale |

`tAge` looks up each clock in the registry and rescales **only** chronological
clocks to age units.

# Reference groups (centring)

All distributed clocks are *relative*: they operate on expression centred
against a reference group.

- **Default (no reference group):** centre on all samples (overall per-gene
  median) — a general cohort readout.
- **Matched controls:** pass `control_group_column` and `control_group_label`
  to `tAge_preprocessing()` to centre on matched controls (recommended for
  treatment-vs-control designs), e.g. `control_group_column = "Genotype",
  control_group_label = "WT"`.

# Session info

```{r session-info}
sessionInfo()
```
