---
title: "Transcriptomic age from bulk RNA-seq with tAge"
author: "Gladyshev Lab"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Transcriptomic age from bulk RNA-seq with tAge}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(tAge)

# This vignette runs on the example data shipped with the package. The
# prediction step additionally needs a Python environment (reticulate +
# joblib/pandas/scikit-learn/numpy) and downloads a small clock model from
# Zenodo. When either is unavailable, the prediction chunks are skipped so the
# vignette still builds. To run the full pipeline locally, set
# RETICULATE_PYTHON to your environment before building:
#   Sys.setenv(RETICULATE_PYTHON = ".venv/bin/python")

.py_ok <- requireNamespace("reticulate", quietly = TRUE) && tryCatch(
  all(vapply(c("joblib", "pandas", "sklearn", "numpy"),
             reticulate::py_module_available, logical(1))),
  error = function(e) FALSE)

# Four clocks: both outcomes x both normalisations, so the figures below have
# something to compare.
.clocks <- NULL
if (.py_ok) {
  .clocks <- tryCatch({
    cl <- list_clocks(type = "EN", species = "Multispecies", tissue = "Multi-Tissue")
    cl <- cl[cl$outcome %in% c("Chronological", "Mortality"), ]
    download_clocks(cl, dest_dir = file.path(tempdir(), "tage-clocks"), quiet = TRUE)
  }, error = function(e) NULL)
}

RUN_MODELS <- .py_ok && !is.null(.clocks) && all(file.exists(.clocks$path))
```

# Overview

`tAge` applies the transcriptomic clocks from
[Tyshkovskiy et al. (2026), *Nature*](https://doi.org/10.1038/s41586-026-10542-3)
to bulk RNA-seq data, predicting **chronological age** (in age units) and
**expected mortality** (as a `log10` hazard ratio).

The workflow: build an `ExpressionSet`, preprocess it, and predict.

# 1. Build an ExpressionSet

We use the mouse example data bundled with the package (genes × samples raw
counts, plus sample metadata).

```{r data}
exprs_data <- load_example_expression_data()
metadata   <- load_example_metadata()

eset <- make_ExpressionSet(exprs_data, metadata, verbose = FALSE)
eset
```

# 2. Preprocess

`tAge_preprocessing()` filters low-expressed genes, maps them to the mouse
Entrez gene space, RLE-normalises, log-transforms, scales (per sample), applies
the YuGene transform, aligns to the clock gene list, and centres the data.

```{r preprocess}
tAge_eset <- tAge_preprocessing(
  eset,
  species = "mouse",
  gene_mapping_type = "Ensembl",
  verbose = FALSE
)
names(tAge_eset)
```

Prediction uses the `scaled_diff` and `yugene_diff` representations.

# 3. Browse and download clocks

The published clocks live on Zenodo; list them from R (reads the bundled
registry):

```{r list-clocks}
head(list_clocks(type = "EN", outcome = "Mortality"))
```

`download_clocks()` fetches a chosen set and returns a `path` column:

```{r download, eval = FALSE}
clocks <- list_clocks(type = "EN", outcome = "Mortality",
                      species = "Multispecies", tissue = "Multi-Tissue")
clocks <- download_clocks(clocks, dest_dir = "clocks")

model_paths <- list(
  scaled_diff = clocks$path[clocks$scaling == "Scaled"],
  yugene_diff = clocks$path[clocks$scaling == "YuGene"]
)
```

# 4. Predict

`predict_tAge()` takes one model per normalisation, so applying several clocks
means one call per outcome. Renaming the columns as we go keeps them apart and
builds the clock table the figures use.

```{r predict, eval = RUN_MODELS}
Sys.setenv(RETICULATE_PYTHON = Sys.getenv("RETICULATE_PYTHON"))  # use your env

predict_one_outcome <- function(outcome) {
  cl <- .clocks[.clocks$outcome == outcome, ]
  res <- predict_tAge(
    tAge_eset,
    model_paths = list(scaled_diff = cl$path[cl$scaling == "Scaled"],
                       yugene_diff = cl$path[cl$scaling == "YuGene"]),
    species = "mouse",
    mode = "EN"
  )
  out <- res[, c("scaled_diff_EN_tAge", "yugene_diff_EN_tAge"), drop = FALSE]
  names(out) <- paste0(outcome, c("_Scaled", "_YuGene"))
  out
}

results <- cbind(
  Biobase::pData(tAge_eset$scaled_diff),
  predict_one_outcome("Chronological"),
  predict_one_outcome("Mortality")
)

# The table the figures use to label rows and pick units per outcome.
clocks_meta <- data.frame(
  filename = c("Chronological_Scaled", "Chronological_YuGene",
               "Mortality_Scaled", "Mortality_YuGene"),
  name     = c("EN Scaled", "EN YuGene", "EN Scaled", "EN YuGene"),
  outcome  = rep(c("Chronological", "Mortality"), each = 2),
  stringsAsFactors = FALSE
)

head(results[, c("Genotype", "Tissue", "Mortality_Scaled")])
```

```{r predict-skipped, echo = FALSE, results = "asis", eval = !RUN_MODELS}
cat("> _Prediction skipped: no Python environment (or no internet to fetch the",
    "models) was available at build time. Set `RETICULATE_PYTHON` and rebuild to",
    "run this step._\n")
```

The mortality values are `log10(hazard ratio)` — small numbers centred near
zero, **not** ages. See below.

# 5. Compare groups

`tage_compare_groups()` is the test behind every figure in the package, and the
same one the TACO web app runs: estimated marginal-mean contrasts of
`tAge ~ group + covariates`, one model per stratum.

```{r stats, eval = RUN_MODELS}
stats <- tage_compare_groups(
  results,
  value_columns   = clocks_meta$filename,
  group_column    = "Genotype",
  reference_group = "WT",
  split_by        = "Tissue",
  p_adjust        = "BH",
  p_adjust_scope  = "across_columns"   # correct across the clocks in one panel
)

stats[, c("value_column", "split", "group2", "estimate",
          "ci_low", "ci_high", "p_adjusted", "label")]
```

`estimate` is always `group2 - group1`, in the clock's own units, and
`ci_low`/`ci_high` are the 95% interval. A few arguments are worth knowing:

- `covariates` puts nuisance variables in the model instead of ignoring them.
- `se_columns` switches to the weighted meta-regression for Bayesian ridge
  clocks — pass the `_sd` columns that `predict_tAge(mode = "BR")` returns.
- `variance_strata` chooses whether the residual variance comes from the two
  compared groups or from every group present.
- `p_adjust_scope` chooses the correction family: `"within_column"` across the
  comparisons of one clock, `"across_columns"` across clocks within a
  comparison, or `"global"`.

# 6. Figures

`tage_clock_forest()` shows every clock at once with its uncertainty. Filled
markers cleared the correction; hollow ones did not.

```{r forest, eval = RUN_MODELS, fig.width = 9, fig.height = 4.5}
tage_clock_forest(
  results,
  clocks_meta     = clocks_meta,
  group_column    = "Genotype",
  reference_group = "WT",
  compare_groups  = "Klotho KO",
  split_by        = "Tissue",
  title           = "Klotho KO vs WT"
)
```

For a single clock, `tage_boxplot()` shows the samples themselves. It annotates
the brackets from the same engine, so the stars agree with the table above.

```{r boxplot, eval = RUN_MODELS, fig.width = 8, fig.height = 4.5}
tage_boxplot(
  results,
  x_var        = "Genotype",
  y_var        = "Mortality_Scaled",
  subgroup_var = "Tissue",
  x_order      = c("WT", "Klotho KO"),
  ylab         = "Mortality tAge, log10 HR"
)
```

```{r figures-skipped, echo = FALSE, results = "asis", eval = !RUN_MODELS}
cat("> _Figures skipped along with the prediction step._\n")
```

When the clocks are module clocks, `tage_module_heatmap()` is the equivalent
figure: modules on the rows, strata on the columns, effects in the cells.

# 7. Interpreting the output

| Outcome | Units | Notes |
|---|---|---|
| Chronological | age (years / months) | normalised age rescaled by species max lifespan |
| Mortality | `log10(hazard ratio)` | **not** an age; higher = higher expected mortality |
| Normalized age | fraction of max lifespan | native scale |

`tAge` looks up each clock in the registry and rescales **only** chronological
clocks to age units.

# 8. Reference groups (centring)

All distributed clocks are *relative*: they operate on expression centred
against a reference group.

- **Default (no reference group):** centre on all samples (overall per-gene
  median) — a general cohort readout.
- **Matched controls:** pass `control_group_column` and `control_group_label`
  to `tAge_preprocessing()` to centre on matched controls (recommended for
  treatment-vs-control designs), e.g. `control_group_column = "Genotype",
  control_group_label = "WT"`.

# Session info

```{r session-info}
sessionInfo()
```
